Online Balanced Graph Partitioning: Optimizations for Large Clusters

### 2020-05-03 paper/disc20clustering.tex
The dead end lower bound (but other interesting comments, too)

\noindent
\textbf{Partitioning model.}
The \emph{balanced repartitioning} problem (\OBRP{})
is a fundamental learning problem
that finds applications in the context of
distributed systems optimization~\cite{repartition-disc}. We are given a set $V$ of $n$ nodes 
(e.g., virtual machines or processes),
initially arbitrarily partitioned into $\ell$~clusters
(e.g., servers or entire racks),
each of size~$k$.
The nodes interact using
a sequence of pairwise communication requests
$\sigma = (u_1,v_1),$ $(u_2,v_2),$ $(u_3,v_3), \ldots$,
where a pair $(u_t,v_t)$ indicates that nodes $u_t$ and $v_t$ exchange a~certain amount of data.
Nodes in $C \subset V$ are \emph{collocated}
if they reside in the same cluster.

@@

\noindent
\textbf{Learning model.}
In addition to the general model, we study a simplified Perfect Partition model (also studied by Henzinger et al.~\cite{sigmetrics19_partitioning} (SIGMETRICS 2019)) that we consider for a stronger lower bound.
There, we assume that $\sigma$
reveals the edges of a static graph
that can be perfectly partitioned
(no external requests are required).

@@

In the paper that originated a study on BRP \cite{repartition-disc} (DISC 2016), authors provided a $O(k^2 \cdot \ell^2)$-competitive algorithm, and a lower bound $\Omega(k)$.
For $k=2$, they presented a $7$-competitive algorithm~\cite{repartition-disc} with substantial ($\Omega(\ell^2)$) additive constant.

The existing results in~\cite{repartition-disc}
and~\cite{sigmetrics19_partitioning}
primarily focus on a model where 
the online algorithm can use augmentation,
i.e., has slightly larger clusters than the offline 
algorithm. In contrast, in our paper, we focus
on a scenario where the nodes need to be
perfectly balanced among the clusters.
The problem has also been studied in a weaker
model where the adversary can only sample
requests from a fixed distribution~\cite{stochastic-ring}.

\maciek{Offline results}
\maciek{Links to paging}

@@

\noindent
\textbf{Ground sets.}
In the lower bound, a~\emph{ground set} is a group of nodes that repeatedly communicate when an algorithm splits them (places at different clusters).
Their role is to ensure that the algorithm maintains a partition of ground sets into clusters.
We emphasize that ground sets are revealed only upon their nodes are split.
\maciek{Other names: division, segmentation, alignment, formation}

@@

\maciek{Eviction - introduce the concept.}

@@

First, we construct a ground set of size $k-1$ on one of the clusters.
In any configuration, there exists a node that accompanies it.
We only issue requests between such accompanying (time-varying) node, and any node that was initially collocated with it.
Repeating such requests, almost every node must visit this cluster.
In comparison, \OPT performs one migration.

\begin{theorem}
	\label{th:lowerbound}
	The competitive ratio of any deterministic online algorithm for \OBRP{} is  $(k \cdot \ell - k - \ell) / 2 = \Omega(k\cdot \ell)$ for any $k\geq 3$ and for any $\ell \geq 2$.
\end{theorem}

\begin{proof}
	\maciek{Better name than $S_1$?}
	Fix any online algorithm \ALG{}.
	Let $I(C)$ denote the cluster where nodes of ground set $C$ are located initially.
	\maciek{A problem with the definition! The first pair does not have the same initial location of its nodes.}
	Initially all nodes are single (they belong to their own trivial ground sets).

@@

	We construct a ground set containing this alone node (and some other node).
	Precisely, let $x_0$  be the alone node from $S_1$ and let $y_0$ be an arbitrary node that does not reside in $S_1$.
	First,
	we construct a ground set $\{x_0, y_0\}$.
	Since \ALG{} has them split, the adversary produces an external request.
	Since $S_1$ contains a ground set of size $k$, \ALG{} must collocate $\{x_0, y_0\}$ in another cluster.
	To preserve a feasible partition of nodes into clusters,
	\ALG must replace $x_0$ with another single node $x_1$.

	We continue to construct ground sets containing the current alone node from $S_1$.
	Precisely, we merge the ground sets of $x_1$ and the largest ground set $C$ s.t.~$I(C) = I(x_1)$.
	To preserve a feasible partition of nodes into clusters,
	\ALG{} must collocate $x_1$ and $C$ in a~cluster other than $S_1$ and
	consequently replace $x_1$ with another alone node $x_2$.

	We end this procedure prematurely (before all nodes visit $S_1$) to keep \OPT's cost low.
	We repeat merging the component of single node $x_i$ on $S_1$ and the largest component $C'$ s.t.~$I(C')=I(x_i)$,
	until there are only two single nodes left that  originate from the same cluster.
	Formally, we denote these two single nodes by $x^*, y^*$, and we have $I(x^*) = I(y^*)$, and~for any other pair of single nodes
	$x'$ and $y'$,
	we have $I(x') \neq I(y')$.
	Note that at this point there are at most $\ell+1$ single nodes left,
	otherwise, there would be more pairs of single nodes that were initially in the same cluster.

	Finally, we bound the cost of \OPT.
	Given this sequence of requests,
	\OPT migrates $\set{x_0,y_0}$ to the cluster $I(x^*)$ by
	swapping $\set{x_0,y_0}$ with $\set{x^*,y^*}$.
	Hence,
	OPT pays for $2$ node swaps and
	\ALG{} incurs at least one swap for each node in the sequence $X := x_0, x_1,\dots$.
	We exclude at most $(k-1) + ( \ell+1)$ nodes out of $k \cdot \ell$ nodes,
	therefore $|X| \geq k \cdot \ell - k - \ell$ \maciek{elaborate. Don't use the word "exclude" unless very obvious. Just sum up the real cost.}.
	Finally, $\ALG/\OPT \geq (k \cdot \ell - k - \ell) / 2\in \Omega(k\cdot\ell)$.
\end{proof}

@@

\section{Improved algorithm for $k=2$}
\label{sec:k2}

@@

\begin{theorem} \label{thm:k=2}
	The algorithm is (strictly) 6-competitive.
\end{theorem}

>>>>

\subsubsection{Partitioning model.}
The \emph{balanced repartitioning} problem (\OBRP{})
is a fundamental learning problem
that finds applications in the context of
distributed systems optimization~\cite{repartition-disc}. We are given a set $V$ of $n$ nodes 
(e.g., virtual machines or processes),
initially arbitrarily partitioned into $\ell$~clusters
(e.g., servers or entire racks),
each of size~$k$.
The nodes interact using
a sequence of pairwise communication requests
$\sigma = (u_1,v_1),$ $(u_2,v_2),$ $(u_3,v_3), \ldots$,
where a pair $(u_t,v_t)$ indicates that nodes $u_t$ and $v_t$ exchange a~certain amount of data.
Nodes in $C \subset V$ are \emph{collocated}
if they reside in the same cluster.

@@

\maciek{The competitive ratio definition and the worst case inputs}

\maciek{Ilustrate the challenges. Rent-or-buy. Decision what to evict}

\subsubsection{Learning model.}
In addition to the general model, we study a simplified Perfect Partition model (also studied by Henzinger et al.~\cite{sigmetrics19_partitioning} (SIGMETRICS 2019)) that we consider for a stronger lower bound.
There, we assume that $\sigma$
reveals the edges of a static graph
that can be perfectly partitioned
(no external requests are required).

@@

The existing results in~\cite{repartition-disc} and~\cite{sigmetrics19_partitioning} primarily focus on a model with resource augmentation, where  the online algorithm has slightly larger clusters than the offline algorithm.
They showed that their lower bound $\Omega(k)$ holds even for a significant resource augmentation.
They provided an algorithm with the competitive ratio independent of $\ell$ using the $(2+\epsilon)$-augmented cluster capacity.

In contrast, we study the non-augmented setting, where the nodes need to be perfectly balanced  among the clusters.
In the paper that originated a study on BRP \cite{repartition-disc} (DISC 2016), authors provided a $O(k^2 \cdot \ell^2)$-competitive algorithm, and a lower bound $\Omega(k)$.
For $k=2$, they presented a $7$-competitive algorithm~\cite{repartition-disc} with a~substantial ($\Omega(\ell^2)$) additive constant.

The problem has also been studied in a weaker
model where the adversary can only sample
requests from a fixed distribution~\cite{stochastic-ring}.

\maciek{Offline results}
\maciek{Links other problems: for $k=2$ a flavor of online matching, for $\ell = 2$ it generalizes paging}

@@

\noindent
\textbf{Ground sets.}
In the lower bound, a~\emph{ground set} is a group of nodes that repeatedly communicate when an algorithm splits them (places at different clusters).
Their role is to ensure that the algorithm maintains a partition of ground sets into clusters.
We emphasize that ground sets are revealed only upon their nodes are split.
%\maciek{Other names: division, segmentation, alignment, formation}

@@

@@

\maciek{Better name than $S_1$? Here and in the proof}
First, we construct a ground set of size $k-1$ on one of the clusters $S_1$.
In any configuration, there exists a node that accompanies it.
We only issue requests between such accompanying (time-varying) node, and any alone node that was initially collocated with it.
Repeating such requests, almost half of all node must visit the cluster $S_1$.
In comparison, \OPT performs one migration.

\begin{theorem}
	\label{th:lowerbound}
	The competitive ratio of any deterministic online algorithm for \OBRP{} is  $\Omega(k\cdot \ell)$ for any $k\geq 3$ and for any $\ell \geq 2$.
\end{theorem}

\begin{proof}
	Fix any online algorithm \ALG{}.
	Assume that $k$ is odd.
	We show how to proceed with even $k$ later in this proof.
	Let $I(z)$ denote the cluster where the node $z$ was located initially.
	Throughout the construction, we maintain ground sets for each node.
	Initially all nodes are \emph{alone} (they belong to their own trivial ground sets).

@@

	We construct a ground set containing this alone node from $S_1$ to force its eviction.
	Precisely, let $x_0$  be the alone node from $S_1$ and let $y_0$ be an arbitrary node that does not reside in $S_1$.
	We construct a ground set $\{x_0, y_0\}$.
	\ALG has them split, and the adversary produces an external request.
	$S_1$ already contains a ground set of size $k-1$, hence \ALG must collocate $\{x_0, y_0\}$ in another cluster.

	We continue to construct ground sets containing the current alone node from $S_1$.
	To preserve a feasible partition of nodes into clusters after collocating $\{x_0, y_0\}$,
	\ALG must replace $x_0$ with another single node (that we name $x_1$).
	Then, we merge the ground sets of $x_1$ and an arbitrary node $y_1$ that was initially collocated with $x_1$.
	Continuing this, we observe a sequence $X := x_0, x_1, x_2\dots$ of alone nodes that visit $S_1$.
	(Note that subsequent ground sets are different than the first ground set $\{x_0, y_0\}$, whose nodes were initially in different clusters.)

	We claim that this procedure can continue as long as a perfect partition of ground sets still exists.
	%and we construct a sequence of ground sets $X := \langle \{x_0, y_0\}, \{x_1, y_2\}, \{x_2, y_2\}, \ldots \rangle$ of length $\Omega(k \cdot \ell)$.
	Consider the moment of creating a ground set $\{x_i, y_i\}$.
	If a perfect partition of ground sets still exists, then
	\ALG{} must collocate $\{x_i,y_i\}$ in a~cluster other than $S_1$.
	To this end, it replaces $x_i$ with another alone node $x_{i+1}$.
	To continue, we must guarantee the existence of an alone node that was initially collocated with $x_{i+1}$.

	We end this procedure prematurely to keep \OPT's cost low.
	The sequence must end when no perfect partition of ground sets exists ---
	however, we end it when \emph{two more} ground sets are still possible to give.
	Precisely, we repeat constructing ground sets,
	until there are only two alone nodes left that originate from the same cluster.
	We denote these two nodes by $x^*, y^*$.
	Observe that at this point there are at most $\ell+1$ alone nodes left:
	otherwise, by pigeon hole principle, there would be more pairs of alone nodes that were initially in the same cluster.

	Now we lower bound the length of the sequence $X := x_0, x_1,\dots$.

	The optimal strategy is to perform two node swaps.
	Precisely, \OPT collocates $\set{x_0,y_0}$ by swapping them with $x^*$ and $y^*$.
	In this configuration, no ground set is split.
	We never issue requests between nodes $x^*, y^*$ that are split in this configuration.

	Finally, we bound the cost of \ALG.
	\ALG incurs at least one swap for each node in the sequence $X := x_0, x_1,\dots$.

	We exclude at most $(k-1) + ( \ell+1)$ nodes out of $k \cdot \ell$ nodes,
	therefore $|X| \geq k \cdot \ell - k - \ell$ \maciek{elaborate. Don't use the word "exclude" unless very obvious. Just sum up the real cost.}.
	Finally, $\ALG/\OPT \geq (k \cdot \ell - k - \ell) / 2\in \Omega(k\cdot\ell)$.
\end{proof}

@@

\section{Improved algorithm for Online Rematching}
\label{sec:k2}

@@

\begin{theorem} \label{thm:k=2}
	The algorithm is (strictly) 6-competitive.
	\maciek{for what lambda?}
\end{theorem}
